\chapter*{Author contributions}
\addcontentsline{toc}{chapter}{Author contributions} % Add the preface to the table of contents as a chapter

This thesis contains two scientific articles, both included here in their original form.

Chapter~\ref{ch:results_particle} contains the article ``Ciliary chemosensitivity is enhanced by cilium geometry and motility'', published in the journal eLife, authored by David J Hickey, Andrej Vilfan, and Ramin Golestanian. The project on which the work was based was conceptualised by AV and RG; the analysis and simulation was designed and implemented by DJH and AV. The manuscript was written by DJH and AV, and edited by DJH, AV and RG. Supervision was contributed by AV and RG.

Chapter~\ref{ch:results_sync} contains the article ``Nonreciprocal interactions give rise to fast cilium synchronisation in finite systems'', currently under review in the journal PNAS, authored by David J Hickey, Ramin Golestanian, and Andrej Vilfan. The project on which the work was based was conceptualised by AV and RG; the analysis and simulation was designed and implemented by DJH and AV. The manuscript was written by DJH and AV, and edited by DJH, AV and RG. Supervision was contributed by AV and RG.

All other work and writing presented in the thesis is entirely the work of DJH. %\todo{come back if the PNAS paper changes state}

\clearpage

\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements} % Add the preface to the table of contents as a chapter

% This thesis summarises work conducted over a period of over three years, beginning in 2019 and ending now in 2023. The research presented here has been published in two different journals\todo{I bloody hope}, and has made its way to several conferences and summer schools in poster and presentation form.

This thesis summarises work conducted over a period of over three years, beginning in 2019 and ending now in 2023. Without the support I have received from others, this work could never have existed, so I would like to take this space to acknowledge their contributions:

I am grateful to my thesis advisory committee members, Prof. Dr. Marcus Müller and Prof. Dr. Annette Zippelius, as well as my supervisors, Prof. Dr. Ramin Golestanian and Dr. Andrej Vilfan, for their constant guidance and support. Of these, I would especially like to thank Andrej Vilfan -- I could not have asked for a better supervisor. These people have been the dynein arms that have been propelling me in the right direction.

I am grateful to everyone who proofread this thesis to make sure it was the best it could be, especially my parents, who have probably read through it more times than I have. Without them, this piece of writing would be like a kidney with no mechanosensitive cilia, i.e. missing some very important parts and being dramatically worse off as a result.

Lastly, if my time spent here has taught me anything, it is that collective behaviour is important at all scales of living systems: just as motile cilia are more efficient when they work in a group, life is much better when your friends have your back. For that reason, I'd like to thank everyone who I've enjoyed spending time with during the last three-and-a-bit years, in Göttingen and elsewhere. You know who you are!

\vspace{1cm}

Best wishes,

David